Abstract Control in Static Analysis

### 2014-09-29 writeup/darais-quals.tex
draft is starting to look good, but still not perfect

\title{Abstract Control in Static Analysis (Draft)}
\author{David Darais}
\date{\today}

@@

The field of static analysis offers many techniques for automatically discovering facts about programs.
For example, a static analysis might be used to verify that a program is memory safe, or that it obeys a privacy policy.
Techniques based on Abstract Interpretation (AI) allow the analysis designer to vary some properties of an analysis.
However, features like context, object, and flow sensitivity must be hard-coded into an analysis based on classic AI.
Abstracting Abstract Machines (AAM) is one design methodology for analysis which exposes more of these features as tuning knobs.
Unfortunately, AAM lacks reusable tools for building analyses, and not all analysis features fit nicely into the framework.

We propose a new compositional framework for designing analyses which supports a tuning knob missing from AAM: abstract control.
We achieve compositionality by refactoring AAM into monad transformers, each of which we relate back to abstract state machines.
When put together, these monad transformers build analyses which are correct by construction.
Furthermore, the monadic abstraction allows us to recover a tuning knob for abstract control when designing analyses.
Abstract control allows one to adjust the flow and path sensitivity of an analysis independent of other design choices.

@@

% Introduction {{{
\section{Introduction}
\label{section:Introduction}

Static analysis is useful for many different situations.
A software developer might use an analysis to find bugs in a program.
A software merchant might use an analysis to detect malware or other security vulnerabilities in software hosted by their app store.
A manufacturer might use an analysis to establish high assurance in the reliability of control software for a car or airplane.
Each of these situations requires a different type of analysis.
To meet the needs of a given problem, an analysis designer has two options: retrofit an existing analysis or design one from scratch.
Retrofitting an existing analysis can be insufficient because the tension between precision and performance is often too high.
Getting an analysis right might require more modification than an existing analysis was designed to support.
Because of this, complex analyses often need to be designed from scratch, and analysis designer need tools to help them with this process.

Designing a static analysis is hard.
Designing \emph{the right} static analysis is even more difficult.
One possible process for arriving at the right static analysis is:
\begin{enumerate}
\item Describe the semantics of your language.
\item Design an analysis which you hope has right properties.
\item Exercise the analysis.
\item If the analysis is too slow or imprecise: repeat from 2.
\end{enumerate}
To get a complex analysis right this process must be repeated multiple times.
Iterating quickly between the design and exercise of an analysis is the key to getting the analysis you wanted without too much pain.

One methodology which aims to ease the design process for static analysis is the Abstracting Abstract Machines (AAM) framework\cite{AAM}.
AAM separates two design steps which are traditionally tightly coupled: describing features of interest and abstracting those features.
Furthermore, AAM shows that the abstraction step is systematic, freeing the analysis designer to spend effort in designing features of interest.
However, AAM doesn't provide reusable tools for the design process; the analysis designer must describe their state machine transition from scratch.
Additionally, some analysis features like flow and path sensitivity don't fit nicely into the AAM framework.

We propose a new compositional framework for designing analyses which supports a tuning knob missing from AAM: abstract control.
To achieve this we refactor AAM into monad transformers.
Using monad transformers, the analysis designer can easily add and remove features from their semantics independently of other features.
The monadic abstraction additionaly exposes a new tuning knob for analyses which we call abstract control.
Abstract control allows one to adjust the flow and path sensitivity of an analysis independent of other design choices.

To justify the soundness of the approach, we show that the monad transformers we use are also galois connection transformers.
This means that any combination of our monad transformers will yield a sound approximation of a state machine semantics.

Our contributions in this work are:
\begin{enumerate}
\item A monadic formulation of AAM which exposes abstract control as a tuning knob.
\item A compositional analysis \emph{implementation} framework for building abstract state machines using monad transformers.
\item A compositional analysis \emph{proof} framework for justifying the correctness of abstract state machines using galois connection transformers.
\item A new monad transformer for nondeterminism which is the key to recovering abstract control as a tuning knob.
\end{enumerate}

The rest of the paper is structured as follows:
Section \ref{Background} covers necessary background material in semantics, analysis, and the CPS language which we use for examples.
Section \ref{MAAM} demonstrates a monadic variant of AAM which supports defining compositional interpreters for analysis.
Section \ref{AbstractControl} discusses abstract control and how we exploit the monadic abstraction to expose a tuning knob for it.
Section \ref{Optimizations} gives two examples of analysis optimizations and how they adapt seamlessly to variations in abstract control.
Section \ref{Correctness} discusses the correctness framework and what key lemmas are involved.
Section \ref{Proofs} shows proof steps in more detail for the arguments made in section \ref{Correctness}.
Sections \ref{RelatedWork} and \ref{FutureWork} discuss relate and future work, and section \ref{Conclusion} concludes.

@@

% Background {{{
\section{Background}
\label{section:Background}
% * Background
%   - Small Step Semantics
%     - CPS Language
%   - Abstract Interpretation (Galois Connections)
%     - Control Flow Analysis
%   - AAM
%   - Monads and Monad Transformers

% Small Step Semantics {{{
\subsection{Small-Step Semantics}
\label{section:Background:SmallStepSemantics}

This work follows the small-step operational approach to language semantics pioneered by Plotkin and Felleisen.
In small-step operational semantics, the meaning of an expression $e$ of a language is given by the transitive closure of some step relation $e ↦ e'$.
This yields a state-machine approach to semantics.
Exploring the meaning of $e$ reduces to exploring all possible transitions reachable from $e$ through $↦$.
This contrasts to previously developed deontational approaches, where the meaning of $e$ must be given all at once by a denotation function $⟦ e ⟧$.
This denotation function can quickly demand lots of mathematical machinery.
Small-step operational methods side-step the need for such machinery.

The small-step operation framework supports arbitrary step relations $e ↦ e'$.  
However for our purposes $↦$ will be always be a function-like form of \emph{decidable} relation, i.e. a multifunction.

@@

% CPS {{{
\subsection{Continuation Passing Style}
\label{section:Background:ContinuationPassingStyle}

We use CPS-IF as an example language to demonstrate the benefits of our framework.
CPS is a syntactically restricted version of the lambda calculus.
CPS-IF is an extension of CPS with conditionals.
We choose CPS because the state machine for the semantics of CPS requires no call stack.
We add conditionals to make things slightly more interesting.

@@

CPS-IF supports integer and boolean literals, addition and subtraction by 1, and testing if an integer is greater than or equal to zero.
The essence of CPS is in the syntactic restriction on application forms $a(a)$ and $a(a,a)$.  
Because $a$ cannot be a nested call expression, evaluating a $Call$ requires no evaluation stack.

All lambda calculus terms can be translated to CPS terms through CPS-conversion.
However, because lambda calculus terms are much easier to read, we will use lambda calculus terms for examples.
We display the CPS-converted version of examples where appropriate.

@@

% Abstract Interpretation {{{
\subsection{Abstract Interpretation}
\label{section:Background:AbstractInterpretation}

Abstract Iterpretation (AI) is a formal framework for program analysis pioneered by Cousot and Cousot.
In the setting of abstract interpretation, a program analysis is just an alternate semantics over an abstract domain.
To distinguish the original semantics from the alternate semantics we call the original semantics the \emph{concrete} semantics.
The alternate semantics is called the \emph{abstract} semantics\footnote{
  This terminology quickly becomes confusing because we will eventually construct an abstract machine for both semantics.
  This yields a concrete abstract machine and an abstract abstract machine. 
  This linguistic pun is exploited mercilessly in the work of Might and Van Horn as they “abstract” abstract abstract machines, yielding abstract abstract abstract machines.
}.

Consider a concrete language $\C$ and a program in that language $(\Ce : \C)$.
In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A relationship between $\C$ and $\A$ that defines when some $(\Ce' : \C)$ and $(\Ae' : \A)$ are related.
  This often takes the form of projecting $\A$ to $\PM(\C)$ and using the subset relation.
\item 
  An abstract version of $\Ce$ our concrete program: $(\Ae : \A)$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A method to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
By convention, we put hats ($\widehat{\;\;}$) on things to name their abstract cousins.
The overall approach is summarized in the following picture:
\input{diagrams/AI.tikz}
This picture takes place at some point in the analysis process.
$\Ce$ is the current program (possibly the result of running the base program for a little while).
$\Ae$ is a valid abstraction of $\Ce$, where this validity is expressed by some relation $R$ holding between $\Ce$ and $\Ae$.
$\Ce'$ is some next state of execution of $\Ce$, and likewise for $\Ae$/$\Ae'$.
In order to be a correct analysis, it must be guaranteed that $\Ce'$ and $\Ae'$ will be related.
The logical structure of the picture, which states the correctness of the analysis, is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce R \Ae) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' R \Ae')
\end{equation*}
The \emph{meaning} of the analysis is entirely subject to the relation $R$ which relates the abstract to the concrete.

@@

% Galois Connections {{{
\subsection{Galois Connections}
\label{section:Background:GaloisConnections}

The AI setting can be elegently simplified and enriched through the use of \emph{galois connections}.
Galois connections serve as a unifying framework for establishing the “relationship between $\C $ and $\A$” mentioned in the previous section.

A galois connection between two posets (sets with a partial order) $\C$ and $\A$ is notated $\C\galois{α}{γ}\A$ and contains:
\begin{itemize}
\item $(α : \C → \A)$ where $α$ is monotonic
\item $(γ : \A → \C )$ where $γ$ is monotonic
\item A proof that $(γ ∘ α)$ is expansive: $∀ (x : \C ), x ⊑ γ(α(x))$
\item A proof that $(α ∘ γ)$ is contractive: $∀ (y : \A), α(γ(y)) ⊑ y$
\end{itemize}
The last two properties can be succinctly stated as $(α ∘ γ ⊑ id ⊑ γ ∘ α)$\footnote{
  This uses the logical monotonicity relation $f ⊑ g ⇔  (x ⊑ y ⇒  f(x) ⊑ g(y))$ for the function space.
}.
Equivalent to all four properties is the property $x ⊑ γ(y) ⇔  α(x) ⊑ y$.

The expansive property corresponds to \emph{soundness}, and the contractive property corresponds to \emph{tightness}.
A sound analysis gives results you can trust.  
A tight analysis promises to be the “best” analysis possible.

\paragraph{Example:}
Given a galois connection $\C\galois{α}{γ}\A$, there exists a galois connection $(\C → \C)\galois{α'}{γ'}(\A → \A)$ where:
\begin{align*}
α'(f : \C → \C) &≔ α ∘ f ∘ γ \\
γ'(g : \A → \A) &≔ γ ∘ g ∘ α \\
\end{align*}

\paragraph{Example:} 
The language $(\PM(ℤ),+,*)$ forms a galois connection with the language $(\PM(\{ EVEN, ODD \}),∧,∨)$ where:
\begin{align*}
α(zs : \PM(ℤ))               &≔ ⋃ \{ \{ EVEN  \;|\; ∃ z ∈ zs ∧ Even(z) \},  \{ ODD   \;|\; ∃ z ∈ zs ∧ Odd(z) \} \} \\
γ(ts : \PM(\{ EVEN, ODD \})) &≔ ⋃ \{ \{ z ∈ ℤ \;|\; EVEN ∈ ts ∧ Even(z) \}, \{ z ∈ ℤ \;|\; ODD ∈ ts ∧ Odd(z) \} \} \\
\end{align*}

Galois connections simplify the AI framework by using $x ⊑ γ(y)$ or (equivalently) $α(x) ⊑ y$ as the relation $(x R y)$.
Galois connections are a natural and general way of placing partial orders \emph{on sets themselves}.
$x ⊑ γ(y)$ can be seen as a heterogenous extension of $x ⊑ y$ when $(x : A)$ and $(y : B)$ live in different sets. 
This heterogenous order is given meaning when there exists a galois connection $A\galois{α}{γ}B$.
One can also think of galois connections as something like an isomorphism, but with a weaker round-trip property.
(An isomorphism would require $α ∘ γ = id = γ ∘ α$.)

Using galois connections, the AI framework introduced in the previous section can be re-stated. In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A galois connection $\C\galois{α}{γ}\A$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A way to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
The overall approach is summarized in the following picture:
\input{diagrams/AIGC.tikz}
As before, the logical structure of the picture is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce ⊑ γ(\Ae)) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' ⊑ γ(\Ae'))
\end{equation*}
The statement of this property can be simplified further as $\Csteps ⊑ γ(\Asteps)$.
This uses the definition of galois connections for function spaces shown in a previous example.

Using a galois connection $\C\galois{α}{γ}\A$ and abstract step function $\Asteps ⊑ γ(\Csteps)$, the analysis story becomes simplified even further:
\begin{itemize}
\item Translate $(\Ce : \C)$ to $(α(\Ce) : \A)$
\item Explore all abstract states $\Ae'$ states reachable from $α(\Ce)$.
\item All reachable concrete states are summarized by projecting each $γ(\Ae')$ back to $\C$.
\end{itemize}

The galois connection framework simultaneously guarantees the \emph{soundness} and \emph{tightness} of this method.
Soundness tells us that if we abstract, take an abstract step, and then concretize, then the result is an approximation of taking a concrete step.
Soundness tells us that we can trust the results of the analysis.
\input{diagrams/GCSoundness.tikz}
Tightness tells us that if we concretize, take a concrete step, and then abstract, then the result must be just as precise as taking an abstract step.
Tightness tells us that our abstract step function isn't losing precision unnecessarily, and that our abstract step $\Asteps$ is provably as strong as possible.
\input{diagrams/GCTightness.tikz}

@@

% Control Flow Analysis {{{
\subsection{Control Flow Analysis}
\label{section:Background:ControlFlowAnalysis}

Control flow analysis is a class of analysis which is particularly important for higher-order languages.
In non-higher-order languages, it is useful to distinguish control-flow from data-flow.
The control flow of the program is a graph of what functions are called from where.
The data flow of a program is a mapping of which values can flow to which variables.
Traditionally one performs a control-flow analysis first, to find out which functions are called, and a data-flow analysis second, using the results of the control flow analysis.

In higher-order languages, data-flow and control-flow are tightly coupled.
Before you can tell which functions will be called, you need to know how values flow, because functions are themselves values.

We note that the distinction between “higher-order” languages and “non-higher-order” languages is a red herring.
Functions can be passed as values in C using function pointers, and object-oriented languages enjoy a similar circularity between control- and data- flow due to method dispatch.
You can read these statements as saying “all languages are higher order, and thus all languages need higher order control flow analysis”.
Or you can read them as saying “all languages are higher order, and we've done just fine without higher oder control flow analysis so far”.
The real distinction is in the code that you end up analysing.

We build on the tradition of control flow analysis (CFA) pioneered by Shivers, as well as recent refinements of CFA developed by Might and Van Horn.
In this tradition, the result of a control flow analysis is closer to what one would think of as a data-flow analysis, only the process accounts for higher order flow.
0CFA, the most basic control flow analysis, merely computes the set of values which might flow to a particular variable.
kCFA is a suite of context sensitive extensions to 0CFA for which functions are analyzed separately for each possible calling context.
We use 0CFA as the example analysis in this paper.
However our framework, and our implementation of it, scale to the full range of context sensitive control flow analyses.

@@

% Monads {{{
\subsection{Monads}
\label{section:Background:Monads}

A basic familiarity with the concept of monads will be necessary for understanding this paper.
Some find it useful to review the concept of functor before learning about monads. 
We adopt this approach for our brief explanation of monads.

A type $(F : Set → Set)$ is called a \emph{functor} if one can:
\begin{itemize}
\item Define $map : ∀ (A, B : Set), (A → B) → (F(A) → F(B))$
\item Prove $map(id) = id$, where $id$ is the identity function $(λ(x) → x)$
\item Prove $map (g ∘ f) = map(g) ∘ map(f)$
\end{itemize}

\paragraph{Example:} 
Lists are functors, where $map$ is defined:
\begin{align*}
    map(f)(xs) ≔ &\case(xs):           \\
          Nil    &→ Nil                \\
         x ∷ xs' &→ f(x) ∷ map(f)(xs') \\
\end{align*}

\paragraph{Example:} $map(isEven)([1, 2, 3, 4]) = [False, True, False, True]$

>>>>

\title{Abstract Control in Static Analysis}
\author{David Darais}
\date{\today}

@@

The field of static analysis offers many techniques for automatically discovering facts about programs.
For example, a static analysis might be used to verify that a program is memory safe, or that it obeys a privacy policy.
Techniques based on Abstract Interpretation (AI) allow the analysis designer to vary some properties of an analysis.
However, features like context, object, and flow sensitivity are often hard-coded into an analysis based on classic AI.
Abstracting Abstract Machines (AAM) is one design methodology for analysis which exposes more of these features as tuning knobs.
Unfortunately, AAM lacks reusable tools for building analyses, and not all analysis features fit nicely into the framework.

We propose a new compositional framework for designing analyses which support a tuning knob missing from AAM: abstract control.
We achieve compositionality by refactoring AAM into monad transformers, each of which we relate back to abstract state machines piecewise.
When put together, these monad transformers build analyses which are correct by construction.
Furthermore, the monadic abstraction allows us to recover a new tuning knob which we call abstract control.
Abstract control allows one to adjust the flow and path sensitivity of an analysis independent of other design choices.

@@

% Introduction {{{
\section{Introduction}
\label{Intro}

Static analysis has many applications.
A software developer might use an analysis to find bugs in a program.
A software merchant might use an analysis to detect malware or other security vulnerabilities in software hosted by their app store.
A manufacturer might use an analysis to establish high assurance in the reliability of control software for a car or airplane.
Each of these situations requires a different type of analysis.
To meet the needs of a given problem, an analysis designer has two options: retrofit an existing analysis or design one from scratch.
Retrofitting an existing analysis can be insufficient because the tension between precision and performance is often too high.
Getting an analysis right might require more modification than an existing analysis was designed to support.
Because of this, complex analyses often need to be designed from scratch, and analysis designer need tools to help them with this process.

Designing a static analysis is hard.
Designing \emph{the right} static analysis is even harder.
One possible process for arriving at the right static analysis is:
\begin{enumerate}
\item Describe the semantics of your language.
\item Design an analysis which you hope has right properties.
\item Exercise the analysis.
\item If the analysis is too slow or imprecise: repeat from 2.
\end{enumerate}
To get a complex analysis right this process must be repeated multiple times.
Iterating quickly between the design and exercise of an analysis is the key to getting the analysis you wanted without too much unnecessary effort.

One methodology which aims to ease the design process for static analysis is the Abstracting Abstract Machines (AAM) 
  framework\cite{van-horn:2010:aam}.
AAM separates two design steps which are traditionally tightly coupled: describing features of interest and abstracting those features.
Furthermore, AAM shows that the abstraction step is systematic, freeing the analysis designer to spend effort in designing features of interest.
However, AAM doesn't provide reusable tools for constructing analyses.
The analysis designer must describe their state machine transition function from scratch.
Additionally, some analysis features like flow and path sensitivity don't fit nicely into the AAM framework.

We propose a new compositional framework for designing analyses which supports a tuning knob missing from AAM: abstract control.
To achieve this we refactor AAM into monad transformers.
Using monad transformers, the analysis designer can easily add and remove features from their semantics independently of other features.
The monadic abstraction additionally exposes a new tuning knob for analyses which we call abstract control.
Abstract control allows one to adjust the flow and path sensitivity of an analysis independent of other design choices.

To justify the soundness of the approach, we show that the monad transformers we use are also Galois connection transformers.
This means that any combination of our monad transformers will yield a sound approximation of a state machine semantics.

Our contributions in this work are:
\begin{enumerate}
\item A monadic formulation of AAM which exposes abstract control as a tuning knob.
\item A compositional analysis \emph{implementation} framework for building abstract state machines using monad transformers.
\item A compositional analysis \emph{proof} framework for justifying the correctness of abstract state machines using Galois connection transformers.
\item A new monad transformer for nondeterminism which serves as our gateway to recovering abstract control as a tuning knob.
\end{enumerate}

The rest of the paper is structured as follows:
Section \ref{Background} covers necessary background material in semantics, analysis, and the CPS language which we use for examples.
Section \ref{CompositionalAnalysis} demonstrates a monadic variant of AAM which supports defining compositional interpreters for analysis.
Section \ref{AbstractControl} discusses abstract control and how we exploit the monadic abstraction to expose a tuning knob for it.
Section \ref{Optimizations} gives two examples of analysis optimizations and how they adapt seamlessly to variations in abstract control.
Section \ref{Correctness} discusses the correctness framework and what key lemmas are involved.
Section \ref{Proofs} shows proof steps in more detail for the arguments made in section \ref{Correctness}.
Sections \ref{RelatedWork} and \ref{FutureWork} discuss relate and future work, and section \ref{Conclusion} concludes.

@@

% Background {{{
\section{Background}
\label{Background}

In this section we discuss background material potentially useful for understanding our compositional analysis framework.
This section provides merely a sketch these concepts, and does not motivate them directly in the context of our work.
Some readers may find it useful to skip ahead to section \ref{CompositionalAnalysis} and chase backlinks as they are needed and motivated.
Or some may prefer to proceed directly with this section, consuming the document in a more bottom-up fashion.
For these readers, we have done our best to keep this section brief.

% Small Step Semantics {{{
\subsection{Small-Step Semantics}
\label{SS-Semantics}

We use the small-step operational approach to semantics pioneered by independently by 
  Plotkin\cite{plotkin:1981:sos} 
  and Felleisen\cite{felleisen:1992:red-sem}.
In small-step operational semantics, expressions in a language are given meaning by mapping them to state machines.
Formally, a small-step semantics for a language $L$ is given by:
\begin{itemize}
\item A state machine with configuration states $\C$.
\item A transition relation $σ \Csteps σ'$ for $(\Ce, \Ce' : \C)$.
\item An embedding function $η : L → \C$.
\end{itemize}
The meaning of an expression $e$ is then given by a final configuration $σ$ reachable by $η(e)$ through $\Csteps$.

Small-step semantics contrast to previously developed denotational approaches.
A denotational semantics for a language $L$ is given by a denotation function $⟦\_⟧$ mapping terms $e$ to objects in a mathematical space.
This denotation function can quickly demand non-trivial mathematical machinery for even very simple languages.
Small-step methods side-step the need for such machinery at the cost of potentially useful mathematical structure.
Many useful analysis properties can be states as properties of execution states of an abstract machine.
Small-step semantics therefore serve as a nice middle-ground marrying formality and simplicity.

@@

% CPS {{{
\subsection{Continuation Passing Style}
\label{CPS}

We use CPS-IF as an example language to demonstrate our framework.
CPS-IF is a variant of continuation passing style λ-calculus extended with conditionals and branching.
We use λ-calculus because it is the simplest language with support for higher-order control flow.
λ-calculus can serve as the core target language for larger languages which appear in practice.
We use CPS because the state machine for the semantics of CPS requires one less component and it simplifies our presentation.
All lambda calculus terms can be translated to CPS terms through CPS-conversion, so there is no loss in language expressivity.
We add conditionals to help exercise analyses which track branching structure in different ways.

@@

CPS-IF supports integer and boolean literals, addition and subtraction by 1, and testing if an integer is greater than or equal to zero.
The essence of CPS is in the syntactic restriction on application forms $a(a)$ and $a(a,a)$.  
Because $a$ cannot be a nested expression, evaluating an $Exp$ requires no evaluation stack.

The state machine $\C$ for CPS-IF is defined as:
\input{snips/CPS-IF-CSigma.tex}
Environments map variables to addresses, and the store maps addresses to values.
Values contain closures, which are lambda expressions paired with their closure environment.

The small-step semantics for CPS-IF are defined in two parts.
First we define a \emph{denotation function} $𝒜 $ for $Atom$ expressions.
\input{snips/CPS-IF-CA.tex}

Second we define a \emph{step relation} (as a function) $ℰ$ for $Exp$ expressions.
\input{snips/CPS-IF-CE.tex}

@@

% Abstract Interpretation {{{
\subsection{Abstract Interpretation}
\label{AI}

Abstract Iterpretation (AI) is a formal framework for program analysis pioneered by Cousot and Cousot\cite{cousot:1977:ai}.
In the setting of abstract interpretation, a program analysis is just an alternate semantics over an abstract domain.
To distinguish the original semantics from the alternate semantics we call the original semantics the \emph{concrete} semantics.
The alternate semantics is called the \emph{abstract} semantics\footnote{
  This terminology quickly becomes confusing when we construct an abstract machine for both semantics.
  This yields a concrete abstract machine and an abstract abstract machine. 
  This linguistic pun is exploited mercilessly in the work of Might and Van Horn as they “abstract” abstract abstract machines.
  We simplify the matter by calling them abstract state machines and concrete state machines.
}.

Consider a concrete state machine $\C$ and a state $(\Ce : \C)$.
In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract state machine $\A$.
\item 
  A relationship between $\C$ and $\A$ that defines when some $(\Ce' : \C)$ and $(\Ae' : \A)$ are related.
  This often takes the form of projecting $\A$ to $\PM(\C)$ and using the subset relation.
\item 
  An abstract version of $\Ce$: $(\Ae : \A)$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A method to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
By convention, we put hats ($\widehat{\;\;}$) on things to name their abstract cousins.
The overall approach is summarized in the following picture:
\input{diagrams/AI.tikz}
This picture takes place at some point in the analysis process.
$\Ce$ is the current state (possibly the result of running the original program for a little while).
$\Ae$ is a valid abstraction of $\Ce$, where this validity is expressed by some relation $R$ holding between $\Ce$ and $\Ae$.
$\Ce'$ is some next state of execution of $\Ce$, and likewise for $\Ae$/$\Ae'$.
In order to be a correct analysis, it must be guaranteed that $\Ce'$ and $\Ae'$ will be related.
The logical structure of the picture, which states the correctness of the analysis $\Asteps$, is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce R \Ae) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' R \Ae')
\end{equation*}
The meaning of the analysis is entirely subject to the relation $R$ which relates the abstract to the concrete.

@@

% Galois Connections {{{
\subsection{Galois Connections}
\label{GaloisConnections}

The AI setting can be elegantly simplified and enriched through the use of \emph{Galois connections}.
Galois connections serve as a unifying framework for establishing the “relationship between $\C $ and $\A$” mentioned in the previous section.

A Galois connection between two posets (sets with a partial order) $\C$ and $\A$ is notated $\C\galois{α}{γ}\A$ and contains:
\begin{itemize}
\item $(α : \C → \A)$ where $α$ is monotonic
\item $(γ : \A → \C )$ where $γ$ is monotonic
\item A proof that $(γ ∘ α)$ is expansive: $∀ (x : \C ), x ⊑ γ(α(x))$
\item A proof that $(α ∘ γ)$ is contractive: $∀ (y : \A), α(γ(y)) ⊑ y$
\end{itemize}
The last two properties can be succinctly stated as $(α ∘ γ ⊑ id ⊑ γ ∘ α)$\footnote{
  This uses the usual monotonicity relation $f ⊑ g ⇔  (x ⊑ y ⇒  f(x) ⊑ g(y))$ for the function space.
}.
Equivalent to all four properties is the property $x ⊑ γ(y) ⇔  α(x) ⊑ y$.

A Galois connection $\C\galois{α}{γ}\A$ can be read “$\A$ is an abstraction of $\C$” and $x ⊑ y$ can be read “$x$ is more precise than $y$”.
The expansive property corresponds to \emph{soundness}, and the contractive property corresponds to \emph{tightness}.
A sound analysis gives results you can trust.  
A tight analysis promises to be the “best” analysis possible.

\paragraph{Example:}
Given a Galois connection $\C\galois{α}{γ}\A$, there exists a Galois connection 
  $(\C \xrightarrow{mon} \C)\galois{α'}{γ'}(\A \xrightarrow{mon} \A)$ where:
\begin{align*}
  α'(f : \C \xrightarrow{mon} \C) &≔ α ∘ f ∘ γ \\
  γ'(g : \A \xrightarrow{mon} \A) &≔ γ ∘ g ∘ α \\
\end{align*}

\paragraph{Example:} 
The language $(\PM(ℤ),+,*)$ forms a Galois connection with the language $(\PM(\{ EVEN, ODD \}),∧,∨)$ where:
\begin{align*}
α(zs : \PM(ℤ))               &≔ \{ EVEN  \;|\; ∃ z ∈ zs ∧ Even(z) \} ∪ \{ ODD   \;|\; ∃ z ∈ zs ∧ Odd(z) \}  \\
γ(ts : \PM(\{ EVEN, ODD \})) &≔ \{ z ∈ ℤ \;|\; EVEN ∈ ts ∧ Even(z) \} ∪ \{ z ∈ ℤ \;|\; ODD ∈ ts ∧ Odd(z) \} \\
\end{align*}

Galois connections simplify the AI framework by using $x ⊑ γ(y)$ or (equivalently) $α(x) ⊑ y$ as the relation $(x R y)$.
Galois connections are a natural and general way of placing partial orders \emph{on sets themselves}.
$x ⊑ γ(y)$ can be seen as a heterogenous extension of $x ⊑ y$ when $(x : A)$ and $(y : B)$ live in different sets. 
One can also think of Galois connections as something like an isomorphism, but with a weaker round-trip property.
(An isomorphism would require $α ∘ γ = id = γ ∘ α$.)

Using Galois connections, the AI framework introduced in the previous section can be re-stated. In the AI framework, an analysis for $\Ce$ is:
\begin{itemize}
\item 
  An abstract language $\A$.
\item 
  A Galois connection $\C\galois{α}{γ}\A$.
\item 
  An abstract version of the $\Csteps$ relation: $(\Asteps : \A × \A → Prop)$
\item 
  A way to explore every state reachable by $\Ae$ under $\Asteps$.
  This often requires $\A$ to be finite.
\end{itemize}
The overall approach is summarized in the following picture:
\input{diagrams/AIGC.tikz}
As before, the logical structure of the picture, which states the correctness of the analysis $\Asteps$, is given by:
\begin{equation*}
∀ (\Ce,\Ce' : \C ) (\Ae,\Ae' : \A), (\Ce ⊑ γ(\Ae)) ∧ (\Ce \Csteps \Ce') ∧ (\Ae \Asteps \Ae') ⇒  (\Ce' ⊑ γ(\Ae'))
\end{equation*}
The statement of this property can be simplified further as $\Csteps ⊑ γ(\Asteps)$.
This uses the definition of Galois connections for function spaces shown in a previous example.

Using a Galois connection $\C\galois{α}{γ}\A$ and abstract step function $\Asteps ⊑ γ(\Csteps)$, an algorithm for analysis can be stated:
\begin{enumerate}
\item Translate $(\Ce : \C)$ to $(α(\Ce) : \A)$
\item Explore all abstract states $\Ae'$ states reachable from $α(\Ce)$.
\end{enumerate}
All reachable concrete states are then summarized by projecting each $γ(\Ae')$ back to $\C$.

The Galois connection framework simultaneously justifies the \emph{soundness} and \emph{tightness} of this method.
Soundness tells us that if we (1) abstract, (2) take an abstract step, and then (3) concretize, 
  then the result is an approximation of (4) taking a concrete step.
Soundness tells us that we can trust the results of the analysis.
\input{diagrams/GCSoundness.tikz}
Tightness tells us that if we (1) concretize, (2) take a concrete step, and then (3) abstract, 
  then the result must be just as precise as (4) taking an abstract step.
Tightness tells us that our abstract step function isn't losing precision unnecessarily, 
  and that our abstract step $\Asteps$ is provably as precise as possible.
\input{diagrams/GCTightness.tikz}
The above diagrams compose to state similar properties about $\Csteps*$ and $\Asteps*$.

@@

% Control Flow Analysis {{{
\subsection{Control Flow Analysis}
\label{CFA}

Control flow analysis is a class of analysis which is particularly important for higher-order languages.
In non-higher-order languages, it can be useful to distinguish control-flow from data-flow.
The control flow of the program is the graph of what functions are called from where.
The data flow of a program is the mapping of which values flow to which variables.
Traditionally one performs a control-flow analysis first to find out which functions are called, 
  and a data-flow analysis second using the results of the control flow analysis.
In higher-order languages, data-flow and control-flow are tightly coupled.
Before you can tell which functions will be called, you need to know how values will flow through the program.

We note that the distinction between “higher-order” languages and “non-higher-order” languages is a red herring.
Functions can be passed as values in C using function pointers.
Object-oriented languages enjoy a similar circularity between control and data flow from dynamic method dispatch.
You can read these statements as saying “all languages are higher order, and thus all languages need higher order control flow analysis”.
Or you can read them as saying “all languages are higher order, and we've done just fine without higher oder control flow analysis so far”.
The distinction in practice comes from the properties you are looking for and the code that you end up analysing.

We build on the tradition of Control Flow Analysis (CFA) pioneered by Shivers\cite{shivers:1991:cfa}.
In this tradition, the result of a control flow analysis is closer to what one would think of as a data-flow analysis.
CFA is nothing more than getting data-flow right in a higher-order setting.
0CFA, the most basic control flow analysis, merely computes the set of values which might flow to a particular variable.
kCFA is a family of context sensitive extensions to 0CFA which separates analysis results for functions based on call site.
We use 0CFA and kCFA as the example analyses in this paper.

@@

% Monads {{{
\subsection{Monads}
\label{Monads}

Some find it useful to review the concept of functor before learning about monads. 
We adopt this approach for our brief explanation of monads.

\paragraph{Functors}
A type $(F : Set → Set)$ is called a \emph{functor} if one can:
\begin{itemize}
\item Define $map : ∀ (A, B : Set), (A → B) → (F(A) → F(B))$
\item Prove $map(id) = id$
\item Prove $map (g ∘ f) = map(g) ∘ map(f)$
\end{itemize}

\begin{example}
Lists are functors, where $map$ is defined:
\begin{align*}
    map(f)(xs) ≔ &\case(xs):           \\
          Nil    &→ Nil                \\
         x ∷ xs' &→ f(x) ∷ map(f)(xs') \\
\end{align*}
\end{example}

\begin{example} 
  $map(isEven)([1, 2, 3, 4]) = [False, True, False, True]$ 
\end{example}

\paragraph{Monads}
A type $(ℳ  : Set → Set)$ is called a \emph{monad} if one can:
\begin{itemize}
\item Define $extend : ∀ (A, B : Set), (A → ℳ (B)) → (ℳ (A) → ℳ (B))$
\item Prove $extend(return) = id$, called the left-unit law.
\item Prove $extend(f)(return(x)) = f(x)$, called the right-unit law.
\item Prove $extend(g)(extend(f)(x)) = extend(λ(y) → extend(g)(f(y)))(x)$ called associativity.
\end{itemize}
The only computational difference from the definition of functor is in the first argument of extend.
For monads, this argument is allowed to be a \emph{monadic} $(A → ℳ (B))$, rather than a pure function $(A → B)$.
